\documentclass[10pt]{article}
\usepackage{fullpage}
\usepackage{longtable}
\newcommand{\okay}{\texttt{[OK]}}
\newcommand{\error}{\texttt{[ERROR]}}
\title{UTK Data Acquisition with Pixie16}
\author{David Miller}
\begin{document}
\maketitle
\setcounter{tocdepth}{1}
\setlength{\LTcapwidth}{\textwidth}
\tableofcontents
\section{Introduction}
This is a guide for the use of the newly revised \textsc{PixieSuite} which is 
the basis for data acquisition with the Pixie 16 Rev. D modules. The suite 
consists of several components: a polling program, tools for configuring pixie, 
and a generalized interface to the Pixie 16 modules. Within the suite, only the 
polling program is specific to the Rev. D modules. However, a recent (later 
than Mar. 29, 2010) distribution of the Pixie API is necessary for proper 
compilation of the programs.
\section{Running an experiment}
\subsection{The basics}
In order to take list mode data files with a stable configuration, the 
\texttt{poll} program is used in combination with \texttt{pacman}. The 
information for the Pixie firmware and DSP parameter configuration is taken 
from a standard \texttt{pixie.cfg} file (see Sec. \ref{sec:pi}) for further 
information) and a \texttt{pxisys.ini} file which is distributed with the Pixie 
software. Starting the polling program is as easy as typing:
\begin{verbatim}
[pixie16@localhost Poll]$ ./poll
\end{verbatim}
Additional options for the program can also be seen by typing:
\begin{verbatim}
[pixie16@localhost Poll]$ ./poll -?
Usage: ./poll [options]
  -f               Fast boot (false by default)
  -h <num>         Dump histogram data every num seconds
  -q, --quiet      Run quietly (false by default)
  --no-wall-clock  Do not insert the wall clock in the data stream
  -r               Display module rates in quiet mode (false by defualt)
  -s <num>         Output statistics data every num seconds
  -t <num>         Sets FIFO read threshold to num\% full (50\% by default)
  -z               Zero clocks on each START_ACQ (false by default)
\end{verbatim}
Once started, one should wait for an appropriate \okay{} for the booting of the 
Pixie 16 modules. This can take a moderate amount of time for a number of 
modules when not using the fast booting option. If an \error{} is given, a 
further descripton is available in the \texttt{Pixie16msg.txt} file.

After the polling program indicates it is waiting for the \textsc{START} 
command, it is ready to receive commands from the HRIBF \texttt{pacman} 
program. In particular, \texttt{poll} will respond to the commands: 
\texttt{startvme}, \texttt{stopvme}, \texttt{initvme}, \texttt{statvme}, 
\texttt{trun bon/boff}, and \texttt{tstop}. These respond in the ``expected'' 
fashion except for \texttt{initvme} which will exit from the \texttt{poll} 
program if the acquisition is currently stopped.

One additional caveat exists when taking data to disk, \texttt{pacman} by
default automatically adds a file marker into the data file for every 10000
records it receives. To accomplish this, it issues a quick \textsc{STOP} and
\textsc{START} command to the front-end machine; therefore, one might observe
occasional ending and restarting of runs in the \texttt{poll} output. This is
no cause for alarm. This can be deactivated by giving \texttt{pacman} the
\texttt{afof} (automatic file-marker off) command.

While the acquisition is running, \texttt{poll} will report on how many network 
chunks into which each FIFO read is divided, and the amount of time (reported 
in microseconds) in between spills (i.e. FIFO reads) as well as the time it 
takes the data to be sent over the network. More information is available when 
not running in quiet mode; however, for high count rates, the constant 
refreshing of the display can cause significant demand on the computer (and 
perhaps some nausea for the user). 

Once finished with the acquisition, issue a \texttt{initvme} from 
\texttt{pacman} to quit or alternatively hit \texttt{Ctrl-C} in the console 
running \texttt{poll}. Either of these will clear up the memory and other 
resources accordingly. For an abnormal exit, a lock file 
(\texttt{/var/lock/PixieInterface}) will remain on the system (see 
Sec.~\ref{sec:pi} for further details).

\subsection{The details}
Prior to being ready to receive commands, several actions are taken by the
polling program. A socket to the \texttt{pacman} computer is established (note
that a separate program \texttt{commtest} exists to test this communication);
the IP adderss of the computer is defined through the environment variable
\texttt{PACHOST} or the local computer if no host is defined. The Pixie modules 
are then initialized and booted, and enough memory is allocated to store data 
from all the FIFOs. The Pixie modules are also told to have their clocks reset 
to zero at the beginning of the next list mode run which is set to be started 
synchronously across all the modules so that the timestamps are properly 
correlated between modules.

\begin{table}[htp]
\label{tab:poll_parameters}
\caption{A few key parameters that one might want to change when setting up the 
polling program, all located near the top of \texttt{poll.cpp}}
\begin{tabular}{lp{3.5in}}
\hline
Variable & Description \\
\hline
\texttt{endRunPause} & The amount of time in microseconds the acquisition waits 
before making the final poll at the end of a run \\ 
\texttt{pollPause} & The amount of time in microseconds between successive 
polls \\
\texttt{pollTries} & While running, the number of times to poll before returning 
to check for \texttt{pacman} commands \\
\texttt{readPause} & The amount of time in microseconds to pause before reading 
the words from a pixie module after polling \\
\texttt{waitPause} & The amount of time to wait before starting to poll when a
partial event is detected \\
\texttt{waitTries} & Number of times to try to read the remainder of a partial
event with an appropriate \texttt{pollPause} delay between attempts \\
\end{tabular}
\end{table}

Once a \textsc{START} command is received, the status of the acquisition is 
properly reset, and a list-mode run is began for all the Pixie modules 
described in the slot definition file (typically \texttt{slot\_def.set}). Note 
that for Rev. D modules, the only mode available is \texttt{LIST\_MODE\_RUN0} 
(i.e. full channel header information with traces), and controlling whether 
traces are read is instead controlled by a \texttt{CHANNEL\_CSRA} flag (see 
Sec.~\ref{sec:setup}).

While running, the pixie cards are then polled for a while (see Table
\ref{tab:poll_parameters}) to see if any module has accumulated enough words
to be read out. Once a module exceeds the threshold, then a short pause is
made and the appropriate number of FIFO words determined in the polling stage
is read out from each module sequentially. Occasionally, the program will read
the FIFO while it is in the process of writing an event resulting in a partial
event in the data stream. To properly account for this, the events are parsed
--- this parsing also allows a quick sanity check of the data based on the
material in the event header. If a partial event exists, \texttt{poll} will
then wait for the remainder of that event to exist in the FIFO (and read out)
before continuing with the reading of the modules. If this wait times out, the
partial event will be moved into the next spill. In the situation where a buffer
is overly full ($>90\%$), the run will stop so the problem can be properly 
addressed; here, partial events will be discarded. Finally, the run status is 
checked for all Pixie modules, and the data is divided into shared memory 
buffers in order to be sent over the network. A quick check of any commands 
from the \texttt{pacman} socket is then done before beginning the polling 
process anew.

After a \textsc{STOP} command is received (or the program is interrupted), the 
acquisition will instruct the modules to end their runs. After waiting a fixed 
amount of time, one final pass is made through the polling process in order to 
extract the last events from the FIFO.

\section{\label{sec:setup}Configuring pixie}
In general, the interface for configuring Rev. D Pixie modules is virtually
identical as that for Rev. A with the commands \texttt{pwrite},
\texttt{pread}, \texttt{pmwrite}, and \texttt{pmread} as well as the
corresponding \texttt{R\_*} and \texttt{W\_*} equivalents. In addition, both
the \texttt{mca} and \texttt{trace} programs as well as their paw versions are
included in this directory. Three new programs join the mix: \texttt{boot},
\texttt{copy\_params}, and \texttt{adjust\_offsets}. Note that \texttt{boot}
must be called at least once before working with the modules; all the other
programs only do a partial booting of the Pixie system. All these programs
(see Table~\ref{tab:setup_programs}) are compiled with the Pixie interface
described in Sec.~\ref{sec:pi} and use the working set file as given in
\texttt{pixie.cfg}. Several paw kumacs also exist for interfacing with the
executables with paw support.
% \begin{table}[htp]
% \begin{tabular}{l p{3in}}
\begin{longtable}{lp{4.5in}}
\caption{\label{tab:setup_programs}List of programs useful for setting up Pixie 
channels. Here, the arguments {\it m,c,p,v} indicate the module, channel, 
parameter, and value of interest respectively. If a given module or channel is 
set to -1, this means all modules or all channels for appropriate programs.} \\
\hline
Program & Description \\
\hline
\endfirsthead
\hline
Program & Description \\
\hline
\endhead 
\texttt{adjust\_offsets} {\it m} & Adjust the \texttt{VOFFSET} parameter for 
each channel such that the baseline is located at \texttt{BASELINEPERCENT} of 
the full ADC range. This is a good starting point but doesn't necessarily 
always work as desired.\\
\texttt{boot} & Boot the Pixie modules with the working set file. \\
\texttt{copy\_params} ... & Behaviour depends on the number of arguments with 
{\it s} and {\it d} referencing the source and destination of the copying 
operation  \\
\hspace{0.2in} {\it sm dm} & \hspace{0.2in}Copy parameters channel by channel 
from one module to another \\
\hspace{0.2in} {\it sm sc dm} & \hspace{0.2in}Copy parameters from one source 
channel to all channels of a destination module \\
\hspace{0.2in} {\it sm sc dm dc} & \hspace{0.2in} Copy parameters from one 
channel to another \\
\texttt{csr\_test} {\it v} & Show the flags set for a given \texttt{CHANNELCSRA} 
value \\
\texttt{find\_tau} {\it m c} & Find the decay constant for a module and channel 
\\
\texttt{mca\_paw} [time] & Start a MCA run storing the result in a paw file 
\texttt{mca.dat} with histogram numbers $100(m+1) + c$ \\
\texttt{PGEN} & \textsc{BASH:} Generate the \texttt{R\_*} and \texttt{W\_*} 
links \\
\texttt{pmread} {\it m p} & Read a module parameter \\
\texttt{pmwrite} {\it m p v} & Write a module parameter \\
\texttt{pread} {\it m c p} & Read a channel parameter \\
\texttt{pwrite} {\it m c p v} & Write a channel parameter \\
\texttt{rate} {\it m c} & Get information about the statistics for the most 
recent run \\
\texttt{RATE} & \textsc{BASH:} \texttt{rate} for the active channel \\
\texttt{R\_*} & \textsc{BASH:} Read a parameter for the active module or 
channel \\
% \texttt{RCHA} & \textsc{BASH:} Dummy script, {\bf do not directly use!} \\
\texttt{READ\_CONFIG} & \textsc{BASH:} Read all the relevant parameters for the 
active module and channel [slowly] \\
% \texttt{RMOD} & \textsc{BASH:} Dummy script, {\bf do not directly use!} \\
\texttt{set\_hybrid} {\it m c} & Set the pileup mode to take full traces for 
piled-up events and only energies for non-piled-up events (Proton catcher 
firmware)\\
\texttt{set\_pileups\_reject} {\it m c} & Set the pileup mode to reject 
piled-up events \\
\texttt{set\_pileups\_only} {\it m c} & Set the pileup mode to only accept 
piled-up events (Proton catcher firmware) \\
\texttt{set\_standard} {\it m c} & Set the pileup mode to standard \\

\texttt{sm} {\it m c} & \textsc{BASH:} Change the active module and channel \\
\texttt{smf} & \textsc{BASH:} Get the active module and channel \\
\texttt{smt} {\it m c v} & \textsc{BASH:} Change the active module and channel 
as well as the default mca time for \texttt{mc.kumac} \\
\texttt{toggle\_catcher} {\it m c} & Toggle the catcher bit of the channel 
CSRA (Proton catcher firmware)\\
\texttt{toggle\_gain} {\it m c} & Toggle the gain bit of the channel CSRA\\
\texttt{toggle\_good} {\it m c} & Toggle the good bit of the channel CSRA\\
\texttt{toggle\_pileup} {\it m c} & Toggle the pileup rejection bit of the 
channel CSRA\\
\texttt{toggle\_polarity} {\it m c} & Toggle the polarity bit of the channel 
CSRA\\
\texttt{toggle\_trace} {\it m c} & Toggle the trace capture bit of the channel 
CSRA\\
\texttt{trace} & Print out mean and deviation of traces for all channels 
acquired by Pixie \\
\texttt{trace\_paw} {\it m c} & As \texttt{trace} but also store the trace in 
a paw file named \texttt{trace.dat} with histogram number $100(m+1) + c$ \\
\texttt{W\_*} & \textsc{BASH:} Write a parameter for the active module or
channel \\
\hline
Paw macros & \\
\hline
\texttt{mca} {\it m c v} & Take a histogramming run for {\it v} seconds and
display the results for a given channel (or entire module) \\
\texttt{mc}  & Take a predestined via \texttt{smt} histogram run \\
\texttt{tra} {\it m c} & Get a trace from a given channel (or entire module)
\\
\texttt{trc} & Get a trace for the channel you have specified via \texttt{sm} \\
% \texttt{WCHA} & \textsc{BASH:} Dummy script, {\bf do not directly use!} \\
% \texttt{WMOD} & \textsc{BASH:} Dummy script, {\bf do not directly use!} \\
\end{longtable}
% \end{tabular}
% \end{table}
A simple \texttt{pixie.cfg} exists so that changes to the firmware and 
configuration files do not require a recompilation of the Pixie interface (see 
Sec.~\ref{sec:pi}).
\section{\label{sec:install}Installing a new copy}
\subsection{Requirements}
Several things are expected to be previously installed on the machine in order 
for the suite to work accordingly.
\begin{itemize}
\item A distribution of Pixie firmware (\texttt{dsp} \& \texttt{firmware} 
directories)
\item Headers and libraries for the Pixie API (\texttt{software} directory)
\item An installed PLX interface and library
\item An appropriate \texttt{pxisys.ini} file for the crate
\item A reasonable set file for configuration
\item A version of PAW from cernlib installed with library 
\texttt{libpacklib\_noshift}
\end{itemize}
\subsection{Installing the PixieSuite}
One aim of the new PixieSuite was to make as many of the relevant 
system-dependent configuration variables located clearly in text files with as 
limited need for recompilation as possible. The easiest way to copy the 
distribution from one place to another is through a tarball which can be 
automatically generated by typing:
\begin{verbatim}
[pixie16@localhost PixieSuite]$ ./makedist.bash
\end{verbatim}
in the top directory which creates a file
\texttt{PixieSuite-Host-DDMMYYYY.tgz} containing the relevant sources (not
binaries) which can be copied to another computer. This command will use the
actual git branch if run from a git repository.

After copying to another computer, one must ensure that the directories to
find the necessary libraries and files are correct. These directories are
included through the files \texttt{makepixie.inc} for those regarding Pixie
and \texttt{makepaw.inc} for those regarding the paw configuration. These use
several values from the bash environment if defined:
\begin{table}[ht]
\label{fig:bashenv}
\caption{Bash environment variables used for setup if defined}
\begin{tabular}{lp{3in}}
\hline
Variable & Description \\
\hline
\texttt{CERNLIB\_GFORTRAN} & Set to some arbitrary value if cernlib has been
compiled with \texttt{gfortran} to properly handle the different name-mangling
scheme \\
\texttt{CERNLIB\_INC}      & Directory containing the header files for the c
ernlib package \\
\texttt{CERNLIB\_LIB}      & Directory containing the libraries for the
cernlib package \\
\texttt{CERNLIB\_NOSHIFT}  & Specify that kernlib and packlib are the noshift
versions \\
\texttt{CERNLIB\_ROOT}     & Sets both the library and header directories 
provided that they are contained in the \texttt{lib} and \texttt{include}
subdirectories respectively \\
\texttt{PACHOST}           & Back-end pacman computer with which \texttt{poll}
communicates \\
\texttt{PLX\_SDK\_DIR}     & Root directory for the PLX driver \\
\texttt{PXI\_ROOT}         & Directory containing the Pixie16 API \\
\end{tabular}
\end{table}
Once these are set up accordingly,
\begin{verbatim}
[pixie16@localhost PixieSuite]$ make
\end{verbatim}
in the top directory should take care of the rest. Of course, the individual 
\texttt{pixie.cfg} files need to be directed as well to where the desired 
firmware and configuration are for the relevant version of the Pixie cards.
\section{\label{sec:pi}The Pixie Interface}
The main purpose of the Pixie Interface is to provide a consistent access and 
response (especially error checking) among all the programs which utilize it. 
Through the interface, access is provided for the common \texttt{Pixie16*} 
functions in the Pixie C API. 

Much of the information about the Pixie setup is given through the use of a 
configuration file (typically \texttt{pixie.cfg}). This is simply a file which 
has a list of lines with a tag follwed by blank space and a value which the 
interface interprets. For example:
\begin{verbatim}
PixieBaseDir      /home/pixie16/Pixie16SoftwareRevD_Ver1.2_08212009
SpFpgaFile        firmware/pixie16s3_fippi_r12281.bin
ComFpgaFile       firmware/syspixie16_revdgeneral.bin
TrigFpgaFile      firmware/pixie16trigger.bin
DspConfFile       dsp/Pixie16DSP.ldr
DspVarFile        dsp/Pixie16DSP.var
DspSetFile        configuration/default.set
DspWorkingSetFile ./default_current.set
ListModeFile      ./listmode.dat
# SlotFile        configuration/slot_def.set
SlotFile          ./slot_def.set
# AltComFpgaFile    firmware/syspixie16_revdgeneral_r20408.bin
# AltDspConfFile    dsp/Pixie16DSP_revd_ornl_she_r22025.ldr
# AltDspVarFile     dsp/Pixie16DSP_revd_ornl_she_r22025.var
# AltSpFpgaFile     firmware/fippixie16_revd_ornl_she_r21208.bin
# AltTrigFpgaFile   firmware/pixie16trigger.bin
# TAG               VALUE

\end{verbatim}
shows a test configuration files. Lines starting with a `\#' are ignored by the 
parser, and all files are taken relative to the \texttt{PixieBaseDir} unless 
they begin with a `.' in which case they are considered relative to the current 
directory. The only other necessary configuration file unique to the interface 
is the \texttt{SlotFile} (though it has the same format as previous programs). 
The expected format is:
\begin{verbatim}
3 Modules
2 Mod 0 
3 Mod 1
4 Mod 2
5 Mod 3
*6 Mod 4
...
\end{verbatim}
Only the first number on the line is considered, starting with the number of
modules and then the slot in which each module is located. Extra numbers at
the end of the file are ignored. This creates a slot map which is used in the
Pixie initialization prior to booting. If a line begins with an asterisk, then
the alternative values of the firmware configuration is used from the
\texttt{pixie.cfg} file.

The Interface also provides a few other (possibly non-portable) functions which 
provide consistency among the other programs. First, multiple interfaces are 
prevented from being opened on the same computer. This prevents hard crashes 
from when multiple processes access the Pixie modules simultaneously. However, 
in the event of an error, the file which locks the interface 
(\texttt{/var/lock/PixieInterface}) might not be removed. The Pixie interface 
then reports which process opened the lock file so that one can verify that the 
process is properly terminated, remove the lock file manually, and start anew.

Two timing functions also provide some semblance of determining (real) elapsed 
time in microseconds: \texttt{usGetTime(ref)} for timing relative to a previous 
reference time and \texttt{usGetDTime()} which gives the timing between 
successive calls to this function.
 
Finally, a colorful display is available for use. This is used by default for 
all programs provided they are ran from an xterm. Hopefully, the contrast of 
bad red versus good green allows for the more rapid identification of where 
troublesome issues appear.
\section{Other programs}
Several other programs that could prove useful are not included in the 
\textsc{PixieSuite}, but probably are located in a shared directory. Of course, 
the pixie scanning code, \texttt{pixie\_ldf\_c}, is necessary for extracting 
information from the data. A new version is available for Rev. D which 
incorporates several (considerable) organizational changes (see separate 
documentation). For the most simple tests, Hui's test programs 
\texttt{Pixie16Boot} and \texttt{Pixie16Test} can shed some light on issues and 
debug problems on the front-end. Two additional programs that can be used to 
work on the ``native'' Pixie formats are \texttt{hexwords} and 
\texttt{read\_events} (Rev. D only). For more complicated configuration of DSP 
variables, the simple setup programs may not suffice and I refer you to either 
\texttt{dsp\_conf} or the python program \texttt{read\_set.py}.

\section{Places for improvement}
Following is a list of possible improvements that could be made (in no 
particular order).
\begin{enumerate}
\item Making the distribution tarball may include extraneous files left in the 
directory by users. This also would theoretically break if the top directory of 
the \textsc{PixieSuite} were the root filesystem.
\item Reading only 1 word from the FIFO is not possible for the Rev. D modules. 
This is implemented in the Pixie Intreface with a rudimentary double buffering 
approach. However, in general, reading of a small number of words introduces a 
significant overhead on the system and a more thorough double buffering approach
might be necessary for maximal throughput.
\item Support for multiple crates will have to be addressed at the time when it 
becomes necessary
\end{enumerate}
\end{document}
